Towards multiversal modal operators for homotopy type theory

### 2024-01-11 slides-hottest2023.lagda.tex
typo

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good.
  \medskip
  \pause

  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
  available in the type~$\NN \to \NN$. Similarly with~\bad{\textsc{dc}}.}}
\end{frame}

@@

  \vspace*{-0.8em}
  For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective''
\end{frame}

@@

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Topos for reifying generic models}
  Grothendieck toposes (= categories of sheaves over sites) are mathematical universes:
  \begin{itemize}
    \item The \hil{generic sequence} is an honest function~$\NN \to X$
    \hil{in} a certain topos~$\mathcal{E}$.
    \item The \hil{generic surjection} is an honest surjection \hil{in} an
    appropriate subtopos of~$\mathcal{E}$.
    \item In the \emph{double-negation subtopos} of the base, a predicate~$Q : \NN
    \to \mathsf{Prop}$ such that
    \[ Q\,x \times Q\,y \longrightarrow x = y \quad\text{and}\quad
      \neg\neg \sum_{n\?\NN} Q\,n \]
    looks like an ordinary natural number.
  \end{itemize}

@@

  \begin{columns}[t]
    \begin{column}{0.80\textwidth}
      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
      that~$2^{\aleph_0} = \aleph_1$.}
      \pause
      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
      \pause
      \ytl{1920s}{Set theorists pursue additional axioms to
      settle~\textsc{ch} \\ (one way or another).}
      \pause
      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+\textsc{ch}.}
      \pause
      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+$\neg$\textsc{ch}.}
      \pause
      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
      the philosophy of set theory.}
      \pause
      %arguing that the program of pursuing
      %additional axioms (while successful in many ways) is doomed to fail
      %to settle~\textsc{ch}.}
      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
      mathematics.}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
    \end{column}

@@

    {\scriptsize
    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010).
    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020). \\
    \par}
    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).

    {\scriptsize
    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021).
    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016). \\
    \par}
    \item (I believe) we have syntactic sheaf models in certain special cases,
    when no coherence issues arise in defining the notion of presheaves.
  \end{itemize}

>>>>

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    Assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \vspace*{-0.6em}
  \emph{Proof.} Let an infinite sequence~$\gamma : \NN \to X \times Y$ be
  given. Write~$\gamma\,k = (\alpha\,k,\beta\,k)$. By the lemma, there is a
  monotonic subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$. Because~$Y$
  is well, there are indices~$n < m$ such that~$\beta\,i_n \leq \beta\,i_m$. As
  also~$\alpha\,i_n \leq \alpha\,i_m$, the sequence~$\gamma$ is good. \qed
  \medskip
  \pause

  \dnote{\emph{We cannot trust~\bad{\textsc{lem}}-provided sequences to be
  available in the type~$\NN \to X$. Similarly with~\bad{\textsc{dc}}.}}
\end{frame}

@@

  \vspace*{-0.8em}
  For instance, for~$P_a\,\sigma \defeq (a \in \sigma)$,
  we have~$\nabla P_a \, \varepsilon$. \quad ``$\genalpha$ is surjective'':
  \AgdaSymbol{(}\AgdaBound{a}\AgdaSpace{}%
  \AgdaSymbol{:}\AgdaSpace{}%
  \AgdaPostulate{X}\AgdaSymbol{)}\AgdaSpace{}%
  \AgdaSymbol{→}\AgdaSpace{}%
  \AgdaDatatype{∇}\AgdaSpace{}%
  \AgdaBound{P$_a$}\AgdaSpace{}%
  \AgdaBound{$\varepsilon$}
\end{frame}

@@

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}
  \pause

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{Topos for reifying generic models}
  Grothendieck toposes (= categories of sheaves over sites) are mathematical universes:
  \begin{itemize}
    \item The \hil{generic sequence} is an honest function~$\NN \to X$
    \hil{in} a certain topos~$\mathcal{E}$.
    \item The \hil{generic surjection} is an honest surjection \hil{in} an
    appropriate subtopos of~$\mathcal{E}$.
    \item In the \emph{double-negation subtopos} of the base, a predicate~$Q : \NN
    \to \mathsf{Prop}$ such that
    \[ Q\,x \times Q\,y \,\longrightarrow\, x = y \quad\text{and}\quad
      \neg\neg \sum_{n\?\NN} Q\,n \]
    looks like an ordinary natural number.
  \end{itemize}

@@

  \begin{columns}[t]
    \begin{column}{0.80\textwidth}
      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
      that~$2^{\aleph_0} = \aleph_1$.}
      \pause
      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
      \pause
      \ytl{1920s}{Set theorists pursue additional axioms to
      settle~\textsc{ch} \\ (one way or another).}
      \pause
      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+\textsc{ch}.}
      \pause
      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+$\neg$\textsc{ch}.}
      \pause
      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
      the philosophy of set theory.}
      \pause
      %arguing that the program of pursuing
      %additional axioms (while successful in many ways) is doomed to fail
      %to settle~\textsc{ch}.}
      \ytl{2016}{Oldenziel proposes to study the modal multiverse of toposes.}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
    \end{column}

@@

    {\scriptsize
    Coquand, Jaber. ``A note on forcing and type theory''. \emph{Fundamenta Informaticae 100} (2010). \\
    Jaber, Lewertowski, Pédrot, Sozeau, Tabareau. ``The definitional side of the forcing''. \emph{Proceedings of LICS '16} (2016). \\
    Pédrot. ``Russian constructivism in a prefascist theory''. \emph{Proceedings of LICS '20} (2020).
    \par}
    \item Given a suitable lex modality, we have a syntactic sheaf model (model of modal types).

    {\scriptsize
    Coquand, Ruch, Sattler. ``Constructive sheaf models of type theory.'' \emph{Math. Struct. Comput. Sci.} 31.9 (2021). \\
    Escardó, Xu. ``Sheaf models of type theory in type theory''. Unpublished (2016). \\
    Quirin. ``Lawvere--Tierney sheafification in Homotopy Type Theory''. PhD thesis (2016).
    \par}
    \item (I believe) we have syntactic sheaf models in certain special cases,
    when no coherence issues arise in defining the notion of presheaves.
  \end{itemize}
